{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2024-12-19 paper.tex
Debugging, concrete programming, design sections

\subsubsection*{Two-Stage Methodology}
Our approach is captured by the interior mode \cite{adam-2021-dsr} of design science research.
To \emph{design} Denicek, we identify a number of \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences (\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and Webnicek (\S\ref{sec:walk}),
a simple web-based end-user programming environment based on the substrate. Although Webnicek can
be used to complete end-user programming tasks, it is optimized for testing the underlying
substrate rather than for usability.

@@

~
% ==================================================================================================

@@

In this section, we use Webnicek, our simple web-based prototype, to illustrate the programming
experiences enabled by the Denicek substrate (\S\ref{sec:impl}). Webnicek is based on a structure
editor that supports navigation in the document and issuing of edit commands. We show two
formative examples (\S\ref{app:examples}) related to conference organization.

@@

@@

Webnicek also accounts for the case where edits recorded in the document are themselves
transformed (when they are reconcilliated with other edits during merging). In this case, Denicek
updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).

@@

Programming by demonstration can be used to define interactive elements in the document. In the
simple case (Fig.~\ref{fig:interactive} (1)), the {\small\Verb_click_} event handler is set to a
reference to a sequence of edits recorded in the document. Clicking the button executes the edits
using the mechanism discussed in \S\ref{sec:impl-pbd}, i.e., by appending them to a history at the
time of saving and merging them with the current history.

The Denicek substrate can be used to generalize the interactions recorded through programming by
demonstration. Our prototype illustrates this option in a limited way. As shown in
Fig.~\ref{fig:interactive} (2), a button to remove all completed TODO items can be generalized
from the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
In addition to the recorded interaction, we manually specify (in the source view) that the edits
should be applied to all elements selected by the {\small\Verb_/items/*_} selector, instead of
the original {\small\Verb_/items/0_} selector (prefix) and that the edits should only be
applied to elements for which the formula (which tests if the checkbox is checked) specified by a
relative selector {\small\Verb_./condition/comp/result_} evaluates to true.

\keyideabox{\faMagic}{Generalization Heuristic}{Specifying generalization manually is cumbersome.
Programming by demonstration systems typically implement heuristic for generalization
\cite{myers-2000-intelligence} infers and suggests such generalizations. When integrated into
a system based on Denicek, such heuristic may also be able to automatically add a formula
based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected items).}

@@

To evalaute formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_}
record into {\small\Verb_<x-evaluated>_} (\circled{2} and \circled{3}), which keeps the previous
formula state in the {\small\Verb_formula_} field and the evaluation result in the {\small\Verb_result_} field.
(Keeping the previous state of the formula is not necessary, but it enables provenance analysis
as discussed in \S\ref{sec:impl-provenance}.) The way edits generated by evaluation are merged
with the document is discussed in \S\ref{sec:impl-incremental}

The counter example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
and programming by demonstration. To implement a counter, the Increment and Decrement buttons
wrap the current counter value in a formula that adds or subtracts 1.

\begin{figure}[t]
\includegraphics[width=0.95\columnwidth,clip,trim=0cm 7cm 8cm 0cm]{fig/counter.pdf}
\caption{Increment wraps the existing count in a formula that adds 1 to the
previous value (1), (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).}
\label{fig:counter}
\vspace{-1em}
\end{figure}

\begin{figure}[t]
\centering
\vspace{-0.5em}
\begin{minipage}{0.6\columnwidth}
  ~\\[1em]
  \includegraphics[width=0.9\columnwidth,clip,trim=0cm 14cm 21cm 0cm]{fig/eval.pdf}
  \end{minipage}%
  \begin{minipage}{0.4\columnwidth}
    \caption{Evaluated edits (A, B) are kept at the top of the history. When they conflict with
    ordinary edits (4), evaluated edits are removed (A).}
    \label{fig:eval}
  \end{minipage}
  \vspace{-1em}
\end{figure}

@@

As illustrated in Fig.~\ref{fig:incremental}, Denicek supports incremental re-computa\-tion.
In the example, the cost of speaker travel depends on the number of speakers, but the cost
of refreshments depends only on two constants. Edit that adds a speaker only invalidates the
former.

The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
Denicek generates \emph{evaluated edits} that are appended to the top of the history.
When subsequent edits are made, they are appended after all non-evaluated edits and the
evaluated edits are pushed through the newly added edits. If the edits conflict (according to
Denicek's conflict detection), affected evaluated edits and edits that depend on them are dropped.

\keyideabox{\faRefresh}{Live Programming}{In the exploratory prototype, Denicek does not
automatically evaluate formulas. This makes it easier to see and understand the evaluation
mechanism, but a realistic system based on the substrate would likely automatically evaluate
formulas to provide a live programming experience \cite{petricek-2020-foundations,rein-2019-live}.}

\begin{figure}[t]
  \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
  \vspace{-0.5em}
  \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
  is added (3), only the results of affected formulas are invalidated.}
  \label{fig:incremental}
\end{figure}

@@

\subsection{Schema Code Co-evolution}
\label{sec:impl-schema}

The problem of schema evolution is that, when the schema used in a database or a program evolves,
data and code that depend on the schema need to evolve correspondingly. The problem is well-known
in database systems \cite{rahm-2006-schema} and has recently been studied in the context of
programming systems \cite{edwards-2025-schema}.

Although Denicek does not explicitly track document structure (schema or type), all documents
have an implicit structure and some of the edits transform the document structure. As
discussed in \S\ref{sec:system-ops}, Denicek automatically updates reference nodes in the
document in this case. This enables a form of schema code co-evolution \cite{edwards-2025-schema}.
The formulas embedded in Denicek documents use reference nodes to refer to both data sources
(in the document) and the results of other computations. Consequently, if the document structure
changes, the formulas are automatically updated.

As noted earier, evaluated edits transform the formula structure (wrapping it in the
{\small\Verb_<x-evaluated>_} record). However, they are marked as non-structural and so
references to formulas are not transformed when the formula is evaluated (otherwise, references
to evaluated formulas would not point to the evaluated result, but to the original unevaluated
definition).

@@

\subsection{Provenance tracking}
\label{sec:impl-provenance}

\subsection{Managed copy \& paste}
\label{sec:impl-copy}

% TODO: Start Implementation section with some overall description of the "formative prototype"
% TODO: In operations - better conflict resolution in terms of effect analysis
% (collect selectors with structure/data effects; merging structure x data effects is OK,
% but data x data or structure x structure is not); - Implement this
% TODO: Add a section after Local-First Editing on the new conflict resolution (with demos)
% TODO: In implementation use effects to better replay recorded edits
% (push through new structural but not data edits)

@@

LISTS vs RECORDS
- are they really different? also MRDT idea - needed for both

PRINCIPLES
* merging of edits should not require looking at current document (non-computationality)
* two ways of pushing edits should be equivalent, i.e.

@@

DIFFICULTIRES
* if you are not able to refer to newly added items, you cannot reasonably do PBD
  => need non-numerical indices

@@

// cf shadow dom v reactu
// something like google forms!
% https://gowthamk.github.io/docs/mrdt.pdf page 11

TAG selectors - sound nice, but they break
(you cannot tell if two selectors overlap e.g. when there is index >< tag)

Although Denicek does not explicitly track document structure (or schema, or type), all documents
have an implicit structure.

c.f. notes

memory mapped graphics etc.

@@

divergence model of challenges paper - requires inverse operation for moving edits backwards

alternative approache is discussed in \S\ref{sec:discuss}
=> PBD - updates the recorded edits this would not be needed if the
history was stored as a graph, akin to ordinary git merging rather than rebasing).

@@

\subsection{Heuristic evaluation}
\subsection{Limitations}
\subsection{Future work}
\label{sec:discuss-future}
use type system to check temporarily invalid state

@@

@@

The design the Denicek substrate, we identified five formative examples shown in
Fig.~\ref{fig:examples}. The examples range from established industry benchmarks
(Todo and Counter apps) to problems posed as schema change challenges \cite{edwards-2025-schema}.
The Denicek substrate then co-evolved with Webnicek, a simple web-based programming environment
built (as directly as possible) on top of the substrate and was used to solve implement the
formative examples.

@@

\begin{tabular}{|l|}
\hline
\\[-1em]
\extablecol{Counter}{\cite{kiss-2014-7guis}}
  {Counter with increment and decrement buttons.}
  {The current count is represented by a formula that is modified by the buttons.\\[-0.2em]\quad\footnotesize
  The user can inspect the evaluation trace to see how the count was modified.}
  {Programming by Demonstration, Incremental Recomputation, End-User Debugging}
\extablecol{Todo}{\cite{osmani-2024-todomvc}}
  {App with controls to add an item and remove all completed.}
  {Adding an item must correctly merge with independently added functionality to \\[-0.2em]\quad\footnotesize
   compute which items are completed and remove them based on a formula result.}
  {Programming by Demonstration, Local-First Collaboration, Incremental Recomputation}
\extablecol{Conference List}{\cite{edwards-2025-schema}}
  {Manage a list of invited conference speakers.}
  {Adding speakers to a list through an in-document user interface merges with \\[-0.2em]\quad\footnotesize
   refactoring that turns the list into a table and separates name from an email.}
  {Local-First Collaboration, Programming by Demonstration}
\extablecol{Conference Budget}{\cite{edwards-2025-schema}}
  {Calculate budget based on a speaker list.}
  {References are updated  when the list is refactored. Only affected formulas\\[-0.2em]\quad\footnotesize
   are recomputed and the user can view elements on which the result depends.}
  {Local-First Collaboration, Incremental Recomputation, End-User Debugging}
\extablecol{Hello World}{\cite{miller-2001-simult}}
  {Normalize the capitalization of two word messages.}
  {An operation to normalize the text in a list item can be recorded and applied to\\[-0.2em]\quad\footnotesize
   all list items or, alternatively, applied directly to all list items.}
  {Programming by Demonstration}
\hline
\end{tabular}

>>>>

\subsubsection*{Two-Stage Methodology}
Our approach is captured by the interior mode \cite{adam-2021-dsr} of design science research.
To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences (\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and Webnicek (\S\ref{sec:walk}),
a simple web-based end-user programming environment based on the substrate. Although Webnicek can
be used to complete end-user programming tasks, it is optimized for testing the underlying
substrate rather than for usability.

@@

% --------------------------------------------------------------------------------------------------

@@

In this section, we use the web-based prototype Webnicek to illustrate programming
experiences enabled by the Denicek substrate (\S\ref{sec:impl}). Webnicek is based on a structure
editor that supports navigation in the document and issuing of edit commands. We follow two
formative examples (\S\ref{app:examples}) related to conference planning.

@@

% --------------------------------------------------------------------------------------------------

@@

Webnicek also accounts for the case where edits recorded in the document are themselves
transformed (when they are reconcilliated with other edits during merging). In this case, Webnicek
updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).

% --------------------------------------------------------------------------------------------------

@@

Programming by demonstration can be used to define interactive elements in the document. In a
simple scenario, illustrated in Fig.~\ref{fig:interactive}~(1), the {\small\Verb_click_} event
handler is set to a reference to a sequence of edits recorded in the document. Clicking the button
executes the edits using the mechanism discussed in \S\ref{sec:impl-pbd}, i.e., Webnicek appends
the edits to a history at the time when the edits were recorded and merges the edits with the
current history.

The use of merging when replaying recorded edits is crucial in both the Todo App example and the
Conference List example (see \S\ref{app:examples}). In both cases, it makes it possible to define
a user interface for adding items (new Todo items, new speakers) and later change the document
structure (refactor speaker list to a table) or add functionality (formula to evaluate whether a
Todo item has been completed) without having to recreate the user interface for adding items.
The use of merging ensures that new items are added in a correct format or with the additional
functionality.

The Denicek substrate can be used to generalize the interactions recorded through programming by
demonstration. Our prototype illustrates this option in a limited way. As shown in
Fig.~\ref{fig:interactive}~(2), a button to remove all completed Todo items can be created by
generalising the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
In addition to the recorded interaction, we manually specify (in the source view) that the edits
should be applied to all elements selected by the {\small\Verb_/items/*_} selector, instead of
the original {\small\Verb_/items/0_} selector (prefix) and that the edits should only be
applied to elements for which the formula (which tests if the checkbox is checked) specified by a
relative selector {\small\Verb_./condition/comp/result_} evaluates to true.

\keyideabox{\faMagic}{Generalization Heuristic}{Specifying generalization manually is cumbersome.
Programming by demonstration systems typically implement heuristic for generalization
\cite{myers-2000-intelligence} infers and suggests such generalizations. If integrated into
a system based on Webnicek, such heuristic could automatically construct a formula
based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected items).}

% --------------------------------------------------------------------------------------------------

@@

To evalaute formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_}
record into {\small\Verb_<x-evaluated>_} (\circled{2} and \circled{3}), which keeps the previous
formula state in the {\small\Verb_formula_} field and the evaluation result in the {\small\Verb_result_} field.
(Keeping the previous state of the formula is not necessary, but it enables provenance analysis
as discussed in \S\ref{sec:impl-provenance}.) The way edits generated by evaluation are merged
with the document is discussed in the next section.

The Counter App example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
and programming by demonstration. To implement a counter, the Increment and Decrement buttons
wrap the current counter value in a formula that adds or subtracts 1.

\keyideabox{\faSuperscript}{Formula Language}{
Webnicek exposes the underlying representation of formulas to the user, but the same representation
can be edited through a user-friendly mechanism such as a block-based editor
\cite{jansen-2019-xlblocks} or a calculation view \cite{sarkar-2018-calc}. The key point
is that Denicek's tree structure makes it easy to embed formulas in document in a uniform
way, edit them and merge them with other changes.}

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[width=0.95\columnwidth,clip,trim=0cm 7cm 8cm 0cm]{fig/counter.pdf}
\vspace{-0.5em}
\caption{Increment wraps the existing count in a formula that adds 1 to the
previous value (1), (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).}
\label{fig:counter}
\vspace{-1em}
\end{figure}

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
\centering
\vspace{-0.5em}
\begin{minipage}{0.55\columnwidth}
  ~\\[1em]
  \includegraphics[width=0.9\columnwidth,clip,trim=0cm 14cm 21cm 0cm]{fig/eval.pdf}
  \end{minipage}%
  \begin{minipage}{0.45\columnwidth}
    \caption{Evaluated edits (A, B) are kept at the top of the history. When they conflict with
    ordinary edits (4), they are removed (A).}
    \label{fig:eval}
  \end{minipage}
  \vspace{-1em}
\end{figure}

% --------------------------------------------------------------------------------------------------

@@

As illustrated in Fig.~\ref{fig:incremental}, Webnicek supports incremental re-computa\-tion.
In the example, the cost of speaker travel depends on the number of speakers, but the cost
of refreshments depends only on two constants. Edit that adds a speaker only invalidates the
former.

The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
Webnicek generates \emph{evaluated edits} that are appended to the top of the history.
When subsequent edits are made, they are appended after all non-evaluated edits and the
evaluated edits are pushed through the newly added edits. If the edits conflict (according to
conflict detection discussed in \S\ref{sec:system-ops}), affected evaluated edits and edits that
depend on them are dropped.

\keyideabox{\faRefresh}{Live Programming}{The Webnicek prototype does not automatically
evaluate formulas. This makes it easier to understand the evaluation
mechanism, but a realistic system based on the substrate could automatically evaluate
formulas to provide a live programming experience \cite{petricek-2020-foundations,rein-2019-live}
and use incremental recomputation for performance~reasons.}

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
  \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
  \vspace{-1em}
  \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
  is added (3), only the results of affected formulas are invalidated.}
  \label{fig:incremental}
  \vspace{-1em}
\end{figure}

% --------------------------------------------------------------------------------------------------

\subsection{Schema Change Control}
\label{sec:impl-schema}

Document structure often needs to evolve \cite{burnett-2014-silos}, as illustrated by our
Conference List example where a list is transformed into a table. When this happens,
data and code that depend on the structure of the document need to evolve correspondingly.
The problem is well-known in database systems \cite{rahm-2006-schema} and has recently been
studied in the context of programming systems \cite{edwards-2025-schema}.

Although Denicek does not explicitly track document structure (schema or type), all documents
have an implicit structure and some edits transform this structure. As
discussed in \S\ref{sec:system-ops}, Denicek automatically updates reference nodes in the
document when edits modify the document structure. This enables a form of schema code
co-evolution \cite{edwards-2025-schema}. Formulas embedded in Denicek documents use reference
nodes to refer to both data sources (in the document) and the results of other computations.
Consequently, if the document structure changes, the formulas are automatically updated.

As noted earier, evaluated edits transform the formula structure (wrapping it in the
{\small\Verb_<x-evaluated>_} record). However, they are marked as non-structural and so
references to formulas are not transformed when the formula is evaluated (otherwise, references
would be updated to point to the original unevaluated formulas).

% --------------------------------------------------------------------------------------------------

@@

% --------------------------------------------------------------------------------------------------

@@

% --------------------------------------------------------------------------------------------------

\subsection{End-User Debugging}
\label{sec:impl-provenance}

The most common kind of end-user programming question is determining whether a value they observe
is right or wrong \cite{kissinger-2006-debugging}. One way to help users answer the question is
to provide an explanation how a value was obtained \cite{ko-2009-whyline}. Webnicek provides a basic
mechanism that highlights document nodes that contributed to a specific computed result,
illustrated in Figure~\ref{fig:provenance}.

The implementation leverages the fact that evaluation keeps the original formula, as discussed
in \S\ref{sec:impl-eval}. When a formula is evaluated, the final {\small\Verb_<x-evaluated>_}
document node contains the result, but also sub-tree that represents a full evaluation trace
\cite{perera-2012-functional}. We analyse the trace, collect all reference nodes in the trace and
highlight all nodes referred to in the computation. Denicek makes this easy as we
only need to collect reference nodes nested in a formula node.

\keyideabox{\faBarChart}{Explanations and Linked Visualizations}{In Webnicek, we implement
provenance analysis to show inputs involved in computation, but information from the execution
trace collected by the Denicek substrate can also be used to provide a detailed
explanation~\cite{perera-2012-functional} or to automatically generate linked visualizations~\cite{perera-2022-linked}.}

% --------------------------------------------------------------------------------------------------

\subsection{Concrete Programming}
\label{sec:impl-copy}

Abstraction is an essential feature of programming, but it has a high cognitive
cost~\cite{blackwell-2002-attention}. Programming by demonstration (\S\ref{sec:impl-pbd})
offers one way of reducing the cost. Another way of making programming more
concrete~\cite{edwards-2004-example,smith-1975-pygmalion} is to support copying of functionality
as in prototype-based object-oriented programming~\cite{randall-1995-self}.

Webnicek supports a functionality akin to managed copy \&
paste~\cite{edwards-2006-copypaste,edwards-2022-copypaste} for formulas. Rather than introducing
abstractions (functions), users can copy and modify formulas to reuse them. However, when the user
discovers an error in the original formula, Webnicek lets them use the merging mechanism to correct
the error in the original formula and all its copies.

The mechanism is shown in Figure~\ref{fig:copypaste}. The user copies an incorrect formula using
the \ident{Copy} edit and then modifies it to use a different data source. They then navigate back
in history to the point before the copying and create a temporary fork of the document. In the
fork, they use \ident{RenameField} to switch the argument of the division. When they merge the
temporary fork into the original document, the \emph{Apply to Newly Added} logic of the merge
operation (\S\ref{sec:system-ops}) duplicates the \ident{RenameField} edits and applies them also
to the copied formula. In other words, the \ident{Copy} edit of the Denicek substrate, alongside
with the fact that formulas are ordinary document nodes, provide keys component for a
straightforward implementation of the managed copy \& paste functionality.

\keyideabox{\faClipboard}{Linked Editing}{Webnicek currently requires users to explicitly
manipulate history to correct error across multiple code clones. Research on managing duplicated
code resulted in multiple tools \cite{toomim-2004-linked,duala-ekoko-2008-clone} with dedicated
user interface to manage clones. The Denicek substrate provides the underlying mechanism that
could be used to implement a more user-friendly interface inspired by those systems.}

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
\vspace{-1em}
\caption{Correcting an incorrect formula. The user uses Copy to create the second
formula (1). They notice an error (2), go back in history to switch left and right argument (3),
merge the change and re-evaluate both formulas (4).}
\label{fig:copypaste}
\vspace{-0.5em}
\end{figure}

@@

\subsection{Design principles}
\label{sec:discuss-principles}

PRINCIPLES
* merging of edits should not require looking at current document (non-computationality)

\subsection{Design considerations}
\label{sec:discuss-considerations}

\subsubsection*{explicit indices}
DIFFICULTIRES
* if you are not able to refer to newly added items, you cannot reasonably do PBD
  => need non-numerical indices

\subsubsection*{mrdts and ordering}
LISTS vs RECORDS
- are they really different? also MRDT idea - needed for both

% https://gowthamk.github.io/docs/mrdt.pdf page 11

@@

\subsubsection*{two-way merging}
divergence model of challenges paper - requires inverse operation for moving edits backwards

\subsubsection*{keeping git tree}

=> PBD - updates the recorded edits this would not be needed if the
history was stored as a graph, akin to ordinary git merging rather than rebasing).

\subsubsection*{Better apply to added}

not so easy because of the non-computationality principle
(we cannot know what will need to happen)

@@

@@

\subsubsection*{Schema vs data}
make schema explicit?

\subsection{Future work}
\label{sec:discuss-future}
Although Denicek does not explicitly track document structure (or schema, or type), all documents
have an implicit structure.
=> use type system to check temporarily invalid state

% memory mapped graphics etc.

@@

\subsection{Heuristic evaluation}
\subsection{Limitations}

RANDOM FORMAL STUFF

@@

DEMOS

// cf shadow dom v reactu
// something like google forms!

* two ways of pushing edits should be equivalent, i.e.

@@

The design the Denicek substrate, we identified six formative examples shown in
Fig.~\ref{fig:examples}. The examples range from established industry benchmarks
(Todo and Counter apps) to cases from literature~\cite{edwards-2022-copypaste} and
problems posed as schema change challenges \cite{edwards-2025-schema}.
The Denicek substrate then co-evolved with Webnicek, a simple web-based programming environment
built (as directly as possible) on top of the substrate and was used to solve implement the
formative examples.

@@

\begin{tabular}{|l|}
\hline
\\[-1em]
\extablecol{Counter App}{\cite{kiss-2014-7guis}}
  {Counter with increment and decrement buttons.}
  {The current count is represented by a formula that is modified by the buttons.\\[-0.2em]\quad\footnotesize
  The user can inspect the evaluation trace to see how the count was modified.}
  {Programming by Demonstration, Incremental Recomputation, End-User Debugging}
\extablecol{Todo App}{\cite{osmani-2024-todomvc}}
  {Buttons to add an item and remove all completed.}
  {Adding an item must correctly merge with independently added functionality to \\[-0.2em]\quad\footnotesize
   compute which items are completed and remove them based on a formula result.}
  {Programming by Demonstration, Local-First Collaboration, Incremental Recomputation}
\extablecol{Conference List}{\cite{edwards-2025-schema}}
  {Manage a list of invited conference speakers.}
  {Adding speakers to a list through an in-document user interface merges with \\[-0.2em]\quad\footnotesize
   refactoring that turns the list into a table and separates name from an email.}
  {Local-First Collaboration, Programming by Demonstration}
\extablecol{Conference Budget}{\cite{edwards-2025-schema}}
  {Calculate budget based on a speaker list.}
  {References are updated  when the list is refactored. Only affected formulas\\[-0.2em]\quad\footnotesize
   are recomputed and the user can view elements on which the result depends.}
  {Local-First Collaboration, Incremental Recomputation, End-User Debugging}
\extablecol{Hello World}{\cite{miller-2001-simult}}
  {Normalize the capitalization of two word messages.}
  {An operation to normalize the text in a list item can be recorded and applied to\\[-0.2em]\quad\footnotesize
   all list items or, alternatively, applied directly to all list items.}
  {Programming by Demonstration}
\extablecol{Traffic Accidents}{\cite{edwards-2022-copypaste}}
  {Compute statistics using two data sources.}
  {Formula to compute statistics can be reused with a different data source;\\[-0.2em]\quad\footnotesize
   error correction is propagated automtically to the copied version of the formula.}
  {Concrete Programming, Incremental Recomputation}
\hline
\end{tabular}
